---
title: "Creating a SpaCET object"
output: rmarkdown::html_vignette
description: >
  Creating a SpaCET object.
vignette: >
  %\VignetteIndexEntry{Creating a SpaCET object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
---
output: github_document
---

This tutorial describes the structure of a `SpaCET` object and the two ways
to create one:

1. automatically from Visium, Visium HD, or Xenium output with
   `create.SpaCET.object.10X`;
2. manually from a prepared count matrix and spatial coordinates with
   `create.SpaCET.object`.

Use the automatic constructor when the original 10x output directory is
available. Use the manual constructor for other platforms or when the
input files have already been processed.

## What is a SpaCET object?

A `SpaCET` object is an S4 object with two slots:

| Slot | Contents |
|---|---|
| `input` | Count matrix, spatial coordinates, optional metadata, image information, platform, and organism |
| `results` | Results added by quality control, deconvolution, spatial correlation, and interaction analyses |

Immediately after object creation, the input data are stored in
`SpaCET_obj@input`, while `SpaCET_obj@results` is usually empty.

The following example uses the original spatial transcriptomics pancreatic
ductal adenocarcinoma dataset from the
[`oldST_PDAC` vignette](oldST_PDAC.html).

```{r inspect-object, eval=FALSE}
library(SpaCET)

oldST_path <- system.file("extdata", "oldST_PDAC", package = "SpaCET")
load(file.path(oldST_path, "st_PDAC.rda"))

SpaCET_obj <- create.SpaCET.object(
  counts = counts,
  spotCoordinates = spotCoordinates,
  metaData = NULL,
  imagePath = NA,
  platform = "oldST",
  organism = "human"
)

str(SpaCET_obj)
```

A newly created object has the following general structure:

``` text
Formal class 'SpaCET' with 2 slots
  ..@ input  :List of 6
  .. ..$ counts         : sparse matrix of genes x spatial locations
  .. ..$ spotCoordinates: data.frame of locations x coordinates
  .. ..$ metaData       : optional data.frame
  .. ..$ image          : list containing the image path and raster
  .. ..$ platform       : character
  .. ..$ organism       : character
  ..@ results: list()
```

## Automatically create a SpaCET object

All supported 10x platforms use the same public constructor:

| Platform and data | Constructor call | Input directory |
|---|---|---|
| Visium | `platform = "Visium"` | Space Ranger output |
| Visium HD, binned | `platform = "VisiumHD"` | One `binned_outputs/square_*um` directory |
| Visium HD, segmented | `platform = "VisiumHD"` | The `segmented_outputs` directory |
| Xenium | `platform = "Xenium"` | Xenium output |

### 10x Visium

The bundled breast cancer dataset is also used in the
[`visium_BC` vignette](visium_BC.html).

```{r automatic-visium, eval=FALSE}
library(SpaCET)

visium_path <- system.file("extdata", "Visium_BC", package = "SpaCET")

SpaCET_obj <- create.SpaCET.object.10X(
  dataPath = visium_path,
  platform = "Visium",
  organism = "human"
)

SpaCET_obj@input$platform
dim(SpaCET_obj@input$counts)
head(SpaCET_obj@input$spotCoordinates)
```

The directory should contain the following core files:

``` text
Visium_output/
├── filtered_feature_bc_matrix/
│   ├── barcodes.tsv.gz
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
├── filtered_feature_bc_matrix.h5
└── spatial/
    ├── scalefactors_json.json
    ├── tissue_positions.csv
    ├── tissue_hires_image.png
    └── tissue_lowres_image.png
```

Either the MEX directory or `filtered_feature_bc_matrix.h5` is sufficient.
Depending on the Space Ranger version, the positions file may instead be
named `tissue_positions_list.csv`.


### 10x Visium HD binned output

Visium HD provides multiple bin sizes. Select one bin size and pass that
directory to the constructor.

```{r automatic-visium-hd-binned, eval=FALSE}
visium_hd_bin_path <- file.path(
  "/path/to/space_ranger/outs",
  "binned_outputs",
  "square_016um"
)

SpaCET_obj <- create.SpaCET.object.10X(
  dataPath = visium_hd_bin_path,
  platform = "VisiumHD",
  organism = "human"
)

SpaCET_obj@input$platform
dim(SpaCET_obj@input$counts)
head(SpaCET_obj@input$spotCoordinates)
```

A selected bin directory should contain:

``` text
square_016um/
├── filtered_feature_bc_matrix/
├── filtered_feature_bc_matrix.h5
└── spatial/
    ├── scalefactors_json.json
    ├── tissue_positions.parquet
    ├── tissue_hires_image.png
    └── tissue_lowres_image.png
```

The returned object uses square bins as spatial observations and records
the platform as `"VisiumHD"`.

### 10x Visium HD segmented output

For cell-segmented Visium HD output, set `dataPath` directly to the
`segmented_outputs` directory.

```{r automatic-visium-hd-segmented, eval=FALSE}
visium_hd_cell_path <- file.path(
  "/path/to/space_ranger/outs",
  "segmented_outputs"
)

SpaCET_obj <- create.SpaCET.object.10X(
  dataPath = visium_hd_cell_path,
  platform = "VisiumHD",
  organism = "human"
)

SpaCET_obj@input$platform
dim(SpaCET_obj@input$counts)
head(SpaCET_obj@input$spotCoordinates)
head(SpaCET_obj@input$metaData)
```

The segmented directory should contain:

``` text
segmented_outputs/
├── cell_segmentations.geojson
├── filtered_feature_cell_matrix/
├── filtered_feature_cell_matrix.h5
└── spatial/
    ├── scalefactors_json.json
    ├── tissue_hires_image.png
    └── tissue_lowres_image.png
```

The constructor reads cell centroids from the segmentation output, matches
them to the cell IDs in `filtered_feature_cell_matrix`, and stores the cell
IDs in `SpaCET_obj@input$metaData$barcode`.

### 10x Xenium

Set `platform = "Xenium"` to read a standard Xenium output directory. The
constructor reads the cell-feature matrix and cell centroids, aligns their
cell IDs, and stores the complete cell table as metadata.

```{r automatic-xenium, eval=FALSE}
xenium_path <- "/path/to/xenium_output"

SpaCET_obj <- create.SpaCET.object.10X(
  dataPath = xenium_path,
  platform = "Xenium",
  organism = "human"
)

SpaCET_obj@input$platform
dim(SpaCET_obj@input$counts)
head(SpaCET_obj@input$spotCoordinates)
head(SpaCET_obj@input$metaData)
```

The Xenium directory should include:

``` text
xenium_output/
├── cell_feature_matrix/
│   ├── barcodes.tsv.gz
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
├── cell_feature_matrix.h5
├── cells.csv.gz
└── cells.parquet
```

Either the MEX directory or `cell_feature_matrix.h5` is sufficient.
Likewise, either `cells.csv.gz` or `cells.parquet` can provide the cell
table. The cell table must contain `cell_id`, `x_centroid`, and
`y_centroid`.

## Manually create a SpaCET object

Use `create.SpaCET.object` when platform output is unavailable, when the
files have already been processed, or when a custom format is being used.
The primary inputs are:

1. `counts`: a gene-by-location count matrix;
2. `spotCoordinates`: a location-by-coordinate matrix or data frame;
3. `imagePath`: the path to a matched tissue image, or `NA`;
4. `platform`: the spatial transcriptomics platform.

`metaData` is optional. When supplied, each row must correspond to one
column of `counts`.

### Original spatial transcriptomics data

The bundled PDAC example stores the count matrix and coordinates in
`st_PDAC.rda`.

```{r manual-oldst, eval=FALSE}
library(SpaCET)

oldST_path <- system.file("extdata", "oldST_PDAC", package = "SpaCET")
load(file.path(oldST_path, "st_PDAC.rda"))

# Genes are rows and spots are columns.
counts[1:6, 1:5]

# Spots are rows and coordinate variables are columns.
head(spotCoordinates)

SpaCET_obj <- create.SpaCET.object(
  counts = counts,
  spotCoordinates = spotCoordinates,
  metaData = NULL,
  imagePath = NA,
  platform = "oldST",
  organism = "human"
)
```

### High-resolution spatial transcriptomics data

The same constructor can be used for high-resolution platforms. The
Slide-seq colorectal cancer example from the
[`hiresST_CRC` vignette](hiresST_CRC.html) supplies a sparse count matrix
and bead coordinates.

```{r manual-hires, eval=FALSE}
library(SpaCET)

hires_path <- system.file("extdata", "hiresST_CRC", package = "SpaCET")
load(file.path(hires_path, "counts.rda"))
load(file.path(hires_path, "spotCoordinates.rda"))

counts[1:6, 1:5]
head(spotCoordinates)

SpaCET_obj <- create.SpaCET.object(
  counts = counts,
  spotCoordinates = spotCoordinates,
  metaData = NULL,
  imagePath = NA,
  platform = "SlideSeq",
  organism = "human"
)
```

### Single-cell-resolution CosMx data

Prepared single-cell-resolution data can be loaded in the same way. The
liver hepatocellular carcinoma CosMx dataset can be downloaded from
[Zenodo](https://zenodo.org/records/15006880/files/LIHC_CosMx_data.rda) or
the
[NIH HPC mirror](https://hpc.nih.gov/~Jiang_Lab/SecAct_Package/LIHC_CosMx_data.rda).
The downloaded R data file contains `counts`, `spotCoordinates`, and
`metaData`.

```{r manual-cosmx, eval=FALSE}
library(SpaCET)

load("LIHC_CosMx_data.rda")

# Count-matrix dimensions: genes x cells.
dim(counts)
#> [1]   1000 460441

spotCoordinates[1:5, ]
#>             coordinate_x_um coordinate_y_um
#> c_2_100_733         6708.16         9033.40
#> c_2_101_240         7500.24         8861.56
#> c_2_101_339         7605.60         8736.04
#> c_2_101_452         7629.84         9172.60
#> c_2_102_179         7868.24         8993.68

metaData[1:5, ]
#>                    cellType         niche
#> c_2_100_733      Tumor_core tumor subtype
#> c_2_101_240 Tumor_interface     interface
#> c_2_101_339      Tumor_core         tumor
#> c_2_101_452      Tumor_core tumor subtype
#> c_2_102_179      Tumor_core         tumor

SpaCET_obj <- create.SpaCET.object(
  counts = counts,
  spotCoordinates = spotCoordinates,
  metaData = metaData,
  imagePath = NA,
  platform = "CosMx",
  organism = "human"
)
```

### Add location-level metadata

Metadata can contain annotations, sample groups, quality-control fields, or
other information associated with spatial locations.

```{r manual-metadata, eval=FALSE}
metaData <- data.frame(
  region = rep(c("tumor", "stroma"), length.out = ncol(counts)),
  row.names = colnames(counts)
)

SpaCET_obj <- create.SpaCET.object(
  counts = counts,
  spotCoordinates = spotCoordinates,
  metaData = metaData,
  imagePath = NA,
  platform = "SlideSeq",
  organism = "human"
)
```

